---
title: "Confidence Intervals Lab 1"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Question 1 (Conceptual)

**Part A:** What is the purpose of a confidence interval?

**Part B:** What does "95\% confidence" actually mean?

**Part C:** What is the formula for the standard error of the sample mean?

**Part D:** What is the formula for a 95\% confidence interval for a population mean?

**Part E:** What type of variability are confidence intervals accounting for?

**Part F:** Do confidence intervals account for biased samples?

---

## Question 2 -- Florida Mercury Levels

This is using the info from a study similar to the Florida lakes fish data we saw in the slides.

Data was collected from 107 lakes in Florida. For each lake, the mercury level (ppm) was computed for a large mouth bass. The average mercury level in the sample was 0.554 ppm.

**Part A:** Describe the parameter in context. What is the corresponding parameter symbol? If we know the value, provide it.

```{r, echo=F, include=F}
# Mu = mean mercury levels of large mouth bass in all florida lakes
```

**Part B:** Describe the statistic in context. State the corresponding statistic symbol. If we know the value, provide it.

```{r, echo=F, include=F}
# x-bar = sample mean mercury level of large mouth bass from 107 florida lakes
```

**Part C:** The standard error for this statistic is 0.051. What is the 95\% confidence interval? (Show your work)

```{r, echo=F, include=F}
0.554 - 2*.051
0.554 + 2*.051
```

**Part D:** Interpret the confidence interval.

```{r, echo=F, include=F}
# We are 95% confident the mean mercury level of large mouth bass in all
# florida lakes is between 0.452 and 0.656 ppm
```

**Part E:** In the U.S., the FDA action level is 1 ppm. Is this safely below the U.S. limit?

```{r, echo=F, include=F}
# Yes. 1 is not within our interval
```

**Part F:** In Canada, the safety limit is 0.5 ppm. Is this clearly above the Canadian limit?

```{r, echo=F, include=F}
# No. 0.5 is within our interval
```

**Part G:** Suppose we didn't know the standard error value in part C, but instead we knew the population standard deviation is 0.20. Create a new 95\% confidence interval using this info. Does your answer to Part F change?

```{r, echo=F, include=F}
0.554 - 2*0.2/sqrt(107)
0.554 + 2*0.2/sqrt(107)
```

**Part H:** What is the practical takeaway from your answer to Part G?