---
title: "Probability Worksheet 2"
output: pdf_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      fig.align = 'center', 
                      fig.width = 4, 
                      fig.height = 4, 
                      message = FALSE, 
                      warning = FALSE)
sol <- FALSE
```

## Rules

For each of these problem, please use notation that we have adopted in class, i.e., events $A$ or $B$, probabilities $P(A)$, expressions $P(A|B)$ or $P(A \text{ or }B)$, etc., in addition to solving them numerically.

## Independence

**Question 1** It is estimated that 9\% of people are left handed. We randomly sample five people from our population. Based on these five, answer the following:

  - What is the probability that all five people are right-handed?
  - What is the probability that all five people are left-handed?
  - What is the probability that that not all five people are right-handed (i.e., probability that *at least one* is left-handed)?
  
\vspace{150pt}

**Question 2** Further assume that sex and handedness are independent, i.e., 

$$
P(\text{right-handed and male}) = P(\text{right-handed}) \times P(\text{Male}).
$$

In a population with equal proportions of men and women, answer the following:

  - What is the probability that the first person is male and right-handed?
  - What is the probability the the first two people are male and right-handed
  - What is the probability that the third person is female and left-handed?
  - What is the probability that the first two people are male and right-handed and the third person is female and left-handed?
  
\newpage

**Question 3** In a standard 52 card deck, assume we draw one card at random. Is the event that a card is a Heart independent of the card being an Ace? Make your argument using the **multiplication rule**



\vspace{75pt}

## Conditional Probability


Below is a table of 6,224 individuals from the year 1721 who were exposed to smallpox in Boston. Doctors at the time believed that exposing a person to the disease in a controlled form (inoculating them) could reduce the likelihood of death.

```{r, echo = FALSE}
library(kableExtra)
library(knitr)
library(openintro)
tt <- table(smallpox)[c(2, 1), c(2,1)] %>% addmargins()
rownames(tt) <- c(rownames(tt)[1:2], "Total")
colnames(tt) <- c(colnames(tt)[1:2], "Total")
tt  %>%  kable() %>% 
  kable_styling(latex_options = "hold_position", full_width = FALSE, font_size = 12) %>% 
  add_header_above(c("", "inoculated" = 2, ""), line_sep = 5)
```

Here, the same table is reproduced, providing joint and marginal probabilities

```{r, echo = FALSE}
library(kableExtra)
library(knitr)
library(openintro)
tt <- table(smallpox)[c(2, 1), c(2,1)] %>% proportions %>%  addmargins()
rownames(tt) <- c(rownames(tt)[1:2], "Total")
colnames(tt) <- c(colnames(tt)[1:2], "Total")
tt  %>%  kable(digits = 4) %>% 
  kable_styling(latex_options = "hold_position", full_width = FALSE, font_size = 12) %>% 
  add_header_above(c("", "inoculated" = 2, ""), line_sep = 5)
```


**Question 1** Write in formal notation the probability that a randomly selected person who was not inoculated died from smallpox and compute the probability

\vspace{75pt}

**Question 2** Determine the probability that an inoculated person died from smallpox. How does this compare with what you found in Question 1?


\vspace{75pt}


**Question 3** Based on your results from Q1 and Q2, does it appear as if inoculation is effective at reducing the risk of smallpox?

\newpage

**Question 4** Let $X$ and $Y$ represent the outcomes of rolling two dice. Answer the following:

 - What is the probability that the first die, $X$, is equal to 1?
 - What is the probability that both $X$ and $Y$ are equal to 1?
 - Use the formula for conditional probability to compute $P(Y = 1| X = 1)$
 - What is $P(Y=1)$? Is this different that what we found in the last part? Explain.
 
 
\vspace{170pt}

**Question 5** Suppose that 80\% of people like peanut butter (obviously), 89\% like jelly, and 78\% like both. Given that a randomly sampled person likes peanut butter, what is the probability that they also like jelly?

\vspace{150pt}

**Question 6** Bob is watching a roulette table in a casino and notices that the last five outcomes landed on black. He figures that the probability of landing on black six times in a row is very small $(1/64)$, so he places a bet on red. What is wrong with this logic? (This is known as the Gambler's Fallacy)


<!-- \newpage -->

<!-- **Birthday Problem** What is the probability that *at least* two people in a room share a birthday? For this problem, we can assume: -->

<!--   - There are 25 individuals in the classroom -->
<!--   - Nobody has a birthday on Feb 29 -->
<!--   - Any individual person is equally likely to be born on any day of the year, i.e., the probability of being born on each day is the same -->
<!--   - Hint -- what does this produce: `prod(1:4)`? -->

<!-- Work with your group to come up with a solution to this problem. Try using the notation we have covered today and Monday to frame this problem in probabilistic statements. You are welcome to use R for this problem  -->