---
title: "Hypothesis Testing Worksheet 3"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Question 1 -- Sleep deprivation, CA vs. OR (2 pts)

A CDC report on sleep deprivation rates shows that the proportion of California residents who reported insufficient rest or sleep during each of the preceding 30 days is 8.0\%, while this proportion is 8.8\% for Oregon residents. These data are based on simple random samples of 11,545 California and 4,691 Oregon residents.

**Goal:** Conduct a hypothesis test to determine if these data provide strong evidence that the rate of sleep deprivation is different for the two states.

**Part A**: What type of hypothesis test is this? (proportion / diff. in proportions/ mean / diff. in means)

```{r}
# diff in proportions
```

**Part B**: Compute $\widehat{p}_{pool}$.

```{r}
(11545*.8 + 4691*.88) / (11545 + 4691)
```

**Part C**: Compute the test-statistic Z.

```{r}
(.88 - .8) / sqrt(.0823*(1-.0823)*(1/11545 + 1/4691))
```

**Part D**: Compute the p-value using `pnorm()`.

```{r}
2*pnorm(16.81, lower.tail=F)
```

**Part E**: Write a conclusion to this question using 'strength of evidence'.

**There is overwhelmingly strong evidence that the two states have different rates of sleep deprivation. Looking at our sample statistics, it looks like Oregon has the higher rate.**

---

# Question 2 -- Find the P-value with t-distribution (.5 pts)

A random sample is selected from an approximately normal population
with an unknown standard deviation. Find the p-value for the given sample size and test statistic using `pt()`.

**Part A**: n = 26, T = 2.485, two-tail test

```{r}
2*pt(2.485, df=25, lower.tail=F)
```


**Part B**: n = 18, T = 0.5, right-tail test

```{r}
pt(.5, df=17, lower.tail=F)
```


---

# Question 3 -- Piano (1 pt)

Georgianna claims that in a small city renowned for its music school, the average child takes less than 5 years of piano lessons. We have a random sample of 35 children from the city, with a sample mean of 4.6 years of piano lessons and a sample standard deviation of 2.2 years.

Evaluate Georgianna’s claim using a hypothesis test.

- state hypotheses
- check conditions
- calculate test-statistic
- p-value
- conclusion

**Solutions:**

$H_0$: $\mu$=5, $H_A$: $\mu <$ 5

There was a random sample of 35 $>$ 30 children, so the conditions are met to use a t-test.

```{r}
# test-statistic
T = (4.6 - 5)/(2.2 / sqrt(35))
T

# p-value
pt(T, df=34)
```

There is very little to no evidence to say that the mean years of piano lessons taken by children is below 5 years.




\newpage

# Question 4 -- Diamonds (1.5 pts)

We have data on two random samples of diamonds: one with diamonds that weigh 0.99 carats and one with diamonds that weigh 1 carat. Each sample has 23 diamonds. Sample statistics for the price per carat of diamonds in each sample are provided below. 

---

### 0.99 Carats

Sample mean = \$44.51, s.d. = \$13.32, n=23

### 1 Carat

Sample mean = \$57.20, s.d. = \$18.19, n=23

---

Assuming that the conditions for conducting inference using the t-distribution are satisfied, perform a hypothesis test to see if there is a difference in population prices per carat of diamonds that weigh 0.99 carats and 1 carat. (Wickham 2016)

**Solutions**:

```{r}
# test-statistic
T = (57.2 - 44.51) / sqrt(13.32^2 / 23 + 18.19^2 / 23)
T

# p-value
pt(T, df=22, lower.tail=F)
```

With a test-statistic of 2.699 and p-value of .007, there is overwhelming evidence the population mean prices of diamonds are different for those that weigh 0.99 carats and those that weigh 1 carat.
