---
title: "Homework 2"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, results='hide')
```

### Directions

For this assignment you should record your answers in an R Markdown file and submit the compiled output as a pdf. At the start of each problem, set the seed to 8.

*Homework #2 is due Friday 9/19 by 10:00pm*
$~$



```{r message = FALSE, warning = FALSE, echo= FALSE}
# load the following packages

```


$~$

### Question #1 

Using a for loop and 1000 iterations, what is the probability of getting at least 10 1s on 20 rolls of a fair 4 sided-die?
```{r, echo=FALSE}
set.seed(8)
x=0
for(i in 1:1000){
  x=x+((sum(sample(c(1:4),20,replace=TRUE)==1))>=10)
}
x

```



$~$

### Question #2

Using the replicate function and 1000 iterations, what is the probability of getting at least 10 1s on 20 rolls of a fair 4 sided-die?
```{r, echo=FALSE}
set.seed(8)
sum(replicate(1000,sum(sample(c(1:4),20,replace=TRUE)==1))>=10)

```



$~$

### Question #3

Write a function that does the following:

Given:

  - seed (integer)
  - number of replications (integer)
  - probability of success (decimal between 0 and 1)
  - number of iterations (integer)
  - number of successes (integer)
  
Returns:

  - probability of getting that at least that number of successes within that number of iterations.

Make sure to name it in a reasonable fashion so you could reuse it later (i.e. do not name it q3_func like I do below)

For example, if we had this function running, we could have run the Lab 2 simulation example (die is 5 or 6, coin is heads) as:

```{r, echo=FALSE}
q3_func=function(seed, numRep,pSucc,numIt,numS){
  set.seed(seed)
  return(sum(replicate(numRep, 
            sum(runif(numIt)<=pSucc))>=numS)/numRep)
}

```

```{r, results='show',echo=TRUE}

q3_func(0, 10000,.65,1000,700)

```

$~$

### Question #4

Demonstrate that your function works by

**Part a** Running it on the Lab 2 simulation example mentioned previously (expected value 6e-04)

**Part b** Running it on the example from Question 2 of this HW (hint: what is the probability of rolling a 1 on a fair 4 sided die?)

$~$

### Question #5

Turn the Question 4 from Homework 1 into a function. Then verify it works on the following test cases. Before running the test cases, how long should each of the vectors be?

```{r,echo=TRUE}
x1 <- c(1,2,3, "four", 5, "six", 7, 6, 2)
x2 <- c(3, 7, "four", 2)
x3 <- c("a","b","c")
x4 <- c(1,2,3,8,9,10)
x5 <- c(4,5,6,7)
x6 <- c(1,2,3,4,5,6,7,8,9,10)
x7 <- c(5.1,4.1,3.9,7.1,6.8,55/10,75/10)
```
